---
title: "APR expected and real behavior"
author: "Carlos Kelly"
date: "22/6/2020"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source('libraries.R')
source('functions.R')
source('file_reading.R')

options(knitr.kable.NA = 0)

```

# Session: Financial needs translation

## Goals

1.- Get to breakeven at June 2021
2.- Maximize our revenue as company
3.- Understand how our unit economics will be in the future

## What are we building?

In order to help Uplift to get to the goals we are building a financial model in Quantrix. 
This shows us the financial position of the company month over month considering multiple dimensions and decisions.

## What are we seeing?

The way that the users are paying the loan is not as the way an amortization table looks. Not only the term, but the credit segment, vertical, and even the vintage the users got their loan appears to be variables that impact on the payment behavior.

In order to create a more realistic financial planning and make decisions to get the first goal done, we need to have a realistic approach to the payment behavior of our costumers. If we use a non-data driven approach (amortization table as it is) the reality is that we could be trying to do things that won't help enough to get us to that goal.

## What do we need?

We need it to be as realistic as possible. Payment behavior, charge off behavior, vintage loss rate, originations, etc..., all of these are very important inputs that allow us to build the financial model.

An input with the realistic payment behavior, across the months on books for each segment, vintage, vertical and term, would help us to have a confident call on when we will get to that goal, and how to do it possible.

This next two things are necessary:
The curves of principal payment (will help us understand the cashflow incomes across the months).
The curves of charge off (will help us to understand how our balance sheet will be month over month).

## What are the curves?

We already have the originations for the future months (actually this could be an input to help us get faster to the goal). We also have the percentage of loan amount that will no get paid (Vintage loss rate). We are assuming that the amount that is getting paid will be the only one to pay interest.

The payment curves represent how much of the 100% of principal paid (loan amount * (1-VLR)) will be paid across the months on books. This doesn't consider down payments, as NLS doesn't show us the amount that is being paid on the first day of loan.
$y_{ijkl} = \frac{principalpaid_{ijkl}}{\sum_{i}^{n}principalpaid_{ijkl}}$
Where i is the month on book, n is the limit of month on books (what I saw is term + 2), j is the credit segment, k is the vertical, and l is the vintage. So practically y is the share of principal paid segmented by credit segment, vertical and vintage

The charge off curves represent how much of the 100% of charge off (loan amount * VLR) will be written off across the months on books. This assumes the same behavior that the NLS is recording on charge off.
The function would be like this
$y_{ijkl} = \frac{chargeoff_{ijkl}}{\sum_{i}^{n}chargeoff_{ijkl}}$

### Principal paid curve

I.e. this loan appears that paid before than his last due date of the amortization table.
```{r amort_table_ex, echo=FALSE, message = FALSE}
set.seed(1234)
sample_paid <-
  loan_tapes[["loan_feed_v1_20200531"]] %>%
  filter(loanstatus == 'Fully Paid') %>%
  sample_n(10)

kable(amort_payments_comparisson(sample_paid[1,]))


```

If we place all the vintages of the 11 term loans having the credit segment as a grouping variable, we can see the principal payment curves among the month on books. 
```{r principal_paid_curves_segment, echo=FALSE, message = FALSE}

  monthly_payments_summary_credit_segment <-
    monthly_payments_summary_fun(max_months_after_term = 2,
                                 credit_segment = credit_segment)
  
  lineplot_principal_paid_share(data = monthly_payments_summary_credit_segment,
                                group_var = credit_segment,
                                term = 11)

```

The same happens with the vertical, it appears that also what the users is asking their loan for impacts on how they behave.
```{r principal_paid_curves_vertical, echo=FALSE, message = FALSE}

  monthly_payments_summary_credit_vertical <-
    monthly_payments_summary_fun(max_months_after_term = 2,
                                 vertical = vertical)
  
  lineplot_principal_paid_share(data = monthly_payments_summary_credit_vertical,
                                group_var = vertical,
                                term = 11)

```


### Charge off curve

The charge off could be explained the same way as the share of principal paid


```{r charge_off_curves, echo=FALSE, message = FALSE}

  monthly_chargeoff_cs_summary <- charge_off_summary_by_segment(term = 11)
  
  monthly_chargeoff_cs_summary %>%
    ggplot(aes(
      x = months_in_books,
      y = chargeoff_principal_share,
      color = as.character(credit_segment)
    )) +
    geom_line() +
    labs(
      title = 'Charge off share among the months in books',
      color = '11 term loans',
      x = 'Months on books',
      y = 'Charge off share'
    ) +
    scale_y_continuous(labels = scales::percent) +
    scale_x_continuous(breaks = round(seq(1, 25, by = 1), 1))

```


# Forecasting the payment behavior {.tabset}

The behavior of the payments of each loan will help us understand how our income will be, given any origination. 
Using the total amount of principal paid as a 100% (this would be seen as the $loanamount(1-VLR)$). Each of the credit segments, verticals, and terms should have an specific behavior for the months on books.

## Behavior of principal payment

We can see that there's a difference between the behavior of payment between 3 and 6 term loans, but it looks like the 6 term is like a stretched 3 term.

```{r principal_paid_graphs_3_6, echo=FALSE, message = FALSE}

monthly_payments_summary <-
  monthly_payments_summary_fun(max_months_after_term = 2)

ggplotly(lineplot_principal_paid_share(data = monthly_payments_summary,
                                       term = 3))

ggplotly(lineplot_principal_paid_share(data = monthly_payments_summary,
                                       term = 6))

```

The behavior of principal paid among the months on books for each term differs, but as the term increases to 11 the first payment takes a higher relative share.

```{r principal_paid_graphs_11, echo=FALSE, message = FALSE}

ggplotly(lineplot_principal_paid_share(data = monthly_payments_summary,
                                       term = 11))

```

As the 18 and 24 term loans are not past due yet, the approach we will use to forecast their behavior is using the 11 term loan and re-sizing it to the given months, something similar to what we saw between the 3 and 6 term loans.
```{r principal_paid_graphs_18_24, echo=FALSE, message = FALSE}

ggplotly(lineplot_principal_paid_share(data = monthly_payments_summary,
                                       term = 18))

ggplotly(lineplot_principal_paid_share(data = monthly_payments_summary,
                                       term = 24))

```


## Data cleanning

In order to create a curve that makes sense we need to clean those months that could have an atypical behavior.

```{r graphs_2, echo=FALSE}

monthly_payments_summary_vintage <-
  monthly_payments_summary_fun(
    only_post_term_vintages = T,
    max_months_after_term = 2,
    vintage = vintage,
    all_dates = T
  )

ggplotly(
  lineplot_principal_paid_share(
    data = monthly_payments_summary_vintage,
    group_var = vintage,
    term = 11
  )
)

```

The previous visualization tells us that there are months that we should not consider, as their behavior has some bumps (i.e. Jan 2017, Feb 2017, May 2017, etc...)


## Interest paid by vintage behavior

The behavior of interest paid is similar to the principal paid, where the first months have the highest share at the first month on book of the loan then starts to decrease.

```{r graphs_interest_paid, echo=FALSE}
    monthly_interest_payments_summary_vintage <-
    monthly_payments_summary_fun(
      only_post_term_vintages = T,
      max_months_after_term = 2,
      vintage = vintage,
      variable_used = interest_paid
    )
  
  plotly::ggplotly(
    lineplot_principal_paid_share(
      data = monthly_interest_payments_summary_vintage,
      group_var = vintage,
      term = 3,
      var = 'Interest'
    )
  )
  
  plotly::ggplotly(
    lineplot_principal_paid_share(
      data = monthly_interest_payments_summary_vintage,
      group_var = vintage,
      term = 6,
      var = 'Interest'
    )
  )
  
  plotly::ggplotly(
    lineplot_principal_paid_share(
      data = monthly_interest_payments_summary_vintage,
      group_var = vintage,
      term = 11,
      var = 'Interest'
    )
  )
  
```

The previous visualization tells us that the 3 and 6 term loans have a constant behavior across the vintages, meanwhile the 11 term loans have some random impact in some of the vintages registered.


## Behavior of customers charge off

Behavior of charge off among the vintages for each term, in order to forecast we shouldn't consider the months with atypical behavior.
Those tend to appear in the vintages that haven't got to past due, and vintages that have been disbursed in the early 2018.   

```{r graphs_charge_off, echo=FALSE}

  plotly::ggplotly(graph_charge_off_vintage(term_filter = 3))
  
  plotly::ggplotly(graph_charge_off_vintage(term_filter = 6))
  
  plotly::ggplotly(graph_charge_off_vintage(term_filter = 11))
  
```

Same as payments behavior, 18 and 24 terms aren't mature enough to be considered for the forecasting model, the approach will be the same, re-sizing 11 term loans charge off.

```{r graphs_charge_off_18_24, echo=FALSE}
  
  plotly::ggplotly(graph_charge_off_vintage(term_filter = 18))
  
  plotly::ggplotly(graph_charge_off_vintage(term_filter = 24))
  
```

# Calculating the APR for Unit Economics {.tabset}

Our Unit economics depend on how the users pay their loan, as the users differ from the amortization table the APR will be modified. There's a need to understand what are the real unit economics and comprehend which of the users have the best possible behavior available to learn which parameters will fit them best, create strategies, give signals when something might brake, etc. Additionally the expected (projected) unit economics helps us understand how our decisions will affect the profitability of the company.

```{r apr_graph, echo=FALSE, message=FALSE}
  
vintage_term_summary <- apr_summary()

graph_comparing_monthly_apr <- graph_apr_loans(vintage_term_summary)

plotly::ggplotly(graph_comparing_monthly_apr)

```

Given the previous paragraph, in this next section we will see how for a given sample of paid loans, the expected, real (charged), and registered (in loan tape) APR differ from each other, and some possible explanations about it.

## Paid loans sample

```{r sampling_paid, echo=FALSE}
  
set.seed(1234)
sample_paid <-
  loan_tapes[["loan_feed_v1_20200531"]] %>%
  filter(loanstatus == 'Fully Paid') %>%
  sample_n(10)

```

Creating a sample of 4 paid loans we will look how the customers in flesh and bones pay their loan

The first loan (`r as.character(sample_paid[1,]$loanid)` loanid) seems that paid before their term completion by `r min(which(sample_paid$principal_paid == 0))` months. 
The interest amount to be paid from the amortization table would have been.
`r sum(amort_payments_comparisson(sample_paid[1,])$amort_interest_paid, na.rm = T)`, 
instead he paid `r sum(amort_payments_comparisson(sample_paid[1,])$interest_paid, na.rm = T)`

Given the APR formula $APR = \frac{12}{term}\frac{\sum(interest)}{principal}$
the amortization table would have an APR of `r round(100*apr_data(sample_paid[1,])$amort, 2)`, 
as with the interest he paid the APR was `r round(100*apr_data(sample_paid[1,])$paid, 2)`, almost the same rate. 
The table we have in the loan tape shows that the APR is `r sample_paid[1,'apr']`

```{r sample_1_paid, echo=FALSE, message = FALSE}

kable(amort_payments_comparisson(sample_paid[1,]))


```


The second loan (`r as.character(sample_paid[2,]$loanid)`) paid according the amortization table. 
The amortization interest was
`r sum(amort_payments_comparisson(sample_paid[2,])$amort_interest_paid, na.rm = T)`, 
he paid `r sum(amort_payments_comparisson(sample_paid[2,])$interest_paid, na.rm = T)`, a few cents above the previous number.

The amortization interest would have an APR of `r round(100*apr_data(sample_paid[2,])$amort, 2)`, 
paid `r round(100*apr_data(sample_paid[2,])$paid, 2)`. 
APR in the loan tape is `r sample_paid[2,'apr']`, 
`r sample_paid[2,'apr'] - round(100*apr_data(sample_paid[2,])$paid, 2)` percentage points higher than the paid APR.

```{r sample_2_paid, echo=FALSE, message = FALSE}

kable(amort_payments_comparisson(sample_paid[2,]))

```

The third (`r as.character(sample_paid[3,]$loanid)`) paid in just 2 payments, in the first and thrid month. 
The amortization interest was 
`r sum(amort_payments_comparisson(sample_paid[3,])$amort_interest_paid, na.rm = T)`, 
instead he paid `r sum(amort_payments_comparisson(sample_paid[3,])$interest_paid, na.rm = T)`

The amortization interest would have an APR of `r round(100*apr_data(sample_paid[3,])$amort, 2)`, 
but paid `r round(100*apr_data(sample_paid[3,])$paid, 2)`. 
APR in loan tape is `r sample_paid[3,'apr']`

```{r sample_3_paid, echo=FALSE, message = FALSE}

kable(amort_payments_comparisson(sample_paid[3,]))

```

The last one (`r as.character(sample_paid[4,]$loanid)`) doesn't have any payments in the database. 
The amortization interest was 
`r sum(amort_payments_comparisson(sample_paid[4,])$amort_interest_paid, na.rm = T)`, 
instead he didn't pay the loan, although it appears as fully paid in the loan tape.

The amortization interest would have an APR of `r round(100*apr_data(sample_paid[4,])$amort, 2)`. 
Loan tape APR is `r sample_paid[4,'apr']`

```{r sample_4_paid, echo=FALSE, message = FALSE}

kable(amort_payments_comparisson(sample_paid[4,]))

```


## Unpaid loans sample

```{r sampling_unpaid, echo=FALSE}
  
set.seed(1234)
sample_unpaid <-
  loan_tapes[["loan_feed_v1_20200531"]] %>%
  filter(!(loanstatus %in% c('Not Originated', 'Cancelled'))) %>%
  sample_n(10)

```

Using the same approach, but now for 4 unpaid loans we will look how the customers behave against the amortization table.

First loan (`r as.character(sample_unpaid[1,]$loanid)` loanid) stop paying in its 5th month.
The interest amount to be paid from the amortization table would have been 
`r sum(amort_payments_comparisson(sample_unpaid[1,])$amort_interest_paid, na.rm = T)`, 
instead he paid `r sum(amort_payments_comparisson(sample_unpaid[1,])$interest_paid, na.rm = T)`

As the APR formula doesn't have any impact on the due loan amount, we will use the same process to calculate it.
the amortization table would have an APR of `r round(100*apr_data(sample_unpaid[1,])$amort, 2)`, 
with the interest he paid the APR was `r round(100*apr_data(sample_unpaid[1,])$paid, 2)`. 
The table we have in the loan tape shows that the APR is `r sample_unpaid[1,'apr']`

```{r sample_1_unpaid, echo=FALSE, message = FALSE}

kable(amort_payments_comparisson(sample_unpaid[1,]))


```


The second loan (`r as.character(sample_paid[2,]$loanid)`) stop paying in its 2nd month. 
The amortization interest was 
`r sum(amort_payments_comparisson(sample_paid[2,])$amort_interest_paid, na.rm = T)`, 
he paid `r sum(amort_payments_comparisson(sample_paid[2,])$interest_paid, na.rm = T)`.

The amortization interest would have an APR of `r round(100*apr_data(sample_paid[2,])$amort, 2)`, 
paid `r round(100*apr_data(sample_paid[2,])$paid, 2)`. 
The table we have in the loan tape shows that the APR is `r sample_paid[2,'apr']`, 
`r sample_paid[2,'apr'] - round(100*apr_data(sample_paid[2,])$paid, 2)` percentage points higher than the paid APR.

```{r sample_2_unpaid, echo=FALSE, message = FALSE}

kable(amort_payments_comparisson(sample_unpaid[2,]))

```

The third (`r as.character(sample_paid[3,]$loanid)`) has the `r as.character(sample_paid[3,]$loanstatus)` status.
But it appears that the loan has been paid.
The amortization interest would be 
`r sum(amort_payments_comparisson(sample_paid[3,])$amort_interest_paid, na.rm = T)`, 
instead he paid `r sum(amort_payments_comparisson(sample_paid[3,])$interest_paid, na.rm = T)`

The amortization interest would have an APR of `r round(100*apr_data(sample_paid[3,])$amort, 2)`, 
but paid `r round(100*apr_data(sample_paid[3,])$paid, 2)`. 
The table we have in the loan tape shows that the APR is `r sample_paid[3,'apr']`

```{r sample_3_unpaid, echo=FALSE, message = FALSE}

kable(amort_payments_comparisson(sample_unpaid[3,]))

```

The last one (`r as.character(sample_paid[4,]$loanid)`) paid according the amortization table. 
Although the status that it has in the loan tape is `r as.character(sample_paid[4,]$loanstatus)`.
The amortization interest would be 
`r sum(amort_payments_comparisson(sample_paid[4,])$amort_interest_paid, na.rm = T)`, he paid `r sum(amort_payments_comparisson(sample_paid[4,])$interest_paid, na.rm = T)`, a few cents more than the amortization table.

The amortization interest would have an APR of `r round(100*apr_data(sample_paid[4,])$amort, 2)`, 
but paid `r round(100*apr_data(sample_paid[4,])$paid, 2)`. 
The table we have in the loan tape shows that the APR is `r sample_paid[4,'apr']`

```{r sample_4_unpaid, echo=FALSE, message = FALSE}

kable(amort_payments_comparisson(sample_unpaid[4,]))

```
